---
title: "Daniel's Notebook"
format: html
---

```{r}
#| label: "libraries"
#| message: false

library(tidyverse)
library(NHANES)
library(car)
library(pROC)
library(mgcv)

```

## Data: NHANES (National Health and Nutrition Examination Survey)

We checked three different data sets:

-   ISLR, Introduction to Statistical Learning (book/package): `wages`

-   National Health and Nutrition Examination Survey: `nhanes`

-   UCI University of California, Irvine, Machine Learning Repository: `AdultUCI`

All three data sets meet the requirements for the project. Given the different analyses we will be conducting in the project, we opted for the data set with the **broadest range of variables**.

And: The topic of health also immediately **appealed to us**.

Selecting a few variables makes handling easier.

```{r}
#| label: "nhanes dataset"
#| include: false

data(NHANES)
# str(NHANES)

# Subset with ~20 variables (all types included)
nhanes_sub <- NHANES %>%
  select(
    # demographie
    SurveyYr, Gender, Age, AgeDecade, Race1, Education, MaritalStatus,
    HHIncome, HHIncomeMid, Poverty, HomeRooms, HomeOwn,
    
    # health (continuous + categorical)
    Height, Weight, BMI, HealthGen,
    Pulse, BPSysAve, BPDiaAve,
    
    # behaviour & life style
    PhysActive, PhysActiveDays, SleepHrsNight, SleepTrouble,
    Alcohol12PlusYr, AlcoholDay, AlcoholYear,
    SmokeNow, Smoke100, SmokeAge,
    
    # mental health
    DaysPhysHlthBad, DaysMentHlthBad,
    LittleInterest, Depressed
  )

str(nhanes_sub)

```

### Types of variables in the nhanes dataset: (i.e. nhanes_sub)

| Variable | Type | Description / Example |
|-------------------|-------------------|----------------------------------|
| **SurveyYr** | categorical | Survey year (`2009_10`, `2011_12`) |
| **Gender** | binary | Gender (`male`, `female`) |
| **Age** | count | Age in years (0–80, integers) |
| **AgeDecade** | categorical | Age in decades (`20-29`, `30-39`, …) |
| **Race1** | categorical | Ethnicity (`Black`, `White`, `Mexican`, `Hispanic`, `Other`) |
| **Education** | categorical | Education level (`8th Grade`, `Some College`, …) |
| **MaritalStatus** | categorical | Marital status (`Married`, `Divorced`, `Widowed`, …) |
| **HHIncome** | categorical | Household income class (`0–4999`, `50–74k`, …) |
| **HHIncomeMid** | continuous | Median household income (numeric, in \$) |
| **Poverty** | continuous | Poverty index (income-to-poverty ratio) |
| **HomeRooms** | count | Number of rooms in the household |
| **HomeOwn** | categorical | Home ownership (`Own`, `Rent`, `Other`) |
| **Height** | continuous | Height in cm |
| **Weight** | continuous | Weight in kg |
| **BMI** | continuous | Body Mass Index |
| **HealthGen** | categorical | Self-rated health (`Excellent`, `Fair`, …) |
| **Pulse** | continuous | Resting heart rate (beats per minute) |
| **BPSysAve** | continuous | Average systolic blood pressure |
| **BPDiaAve** | continuous | Average diastolic blood pressure |
| **PhysActive** | binary | Physically active (`Yes`/`No`) |
| **PhysActiveDays** | count | Number of physically active days per week (0–7) |
| **SleepHrsNight** | count | Hours of sleep per night |
| **SleepTrouble** | binary | Trouble sleeping (`Yes`/`No`) |
| **Alcohol12PlusYr** | binary | Ever drank alcohol (≥12 drinks per year, `Yes`/`No`) |
| **AlcoholDay** | count | Number of alcoholic drinks per day |
| **AlcoholYear** | count | Number of alcoholic drinks per year |
| **SmokeNow** | binary | Currently smokes (`Yes`/`No`) |
| **Smoke100** | binary | Ever smoked ≥100 cigarettes (`Yes`/`No`) |
| **SmokeAge** | count | Age at smoking initiation |
| **DaysPhysHlthBad** | count | Days with poor physical health (0–30) |
| **DaysMentHlthBad** | count | Days with poor mental health (0–30) |
| **LittleInterest** | binary | Little interest in activities (`Yes`/`No`) |
| **Depressed** | categorical | Times feeling depressed (None, Most, Several, NAs) |

## Binomial models

### Make outcome variable Depressed (three levels) binary

```{r}
#| label: "Depressed binomial"

summary(nhanes_sub$Depressed)

nhanes_sub <- nhanes_sub %>%
  mutate(
    Depressed_bin = case_when(
      Depressed == "None" ~ "no",
      Depressed %in% c("Several", "Most") ~ "yes",
      TRUE ~ NA_character_
    ),
    Depressed_bin = factor(Depressed_bin, levels = c("no", "yes"))
  )

# Check the recode
table(nhanes_sub$Depressed, nhanes_sub$Depressed_bin, useNA = "ifany")

```

### Visual inspection of Variable Depressed_bin

```{r}
#| label: "Depressed_bin Visuals"

# Age vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = Depressed_bin, y = Age)) +
  geom_boxplot()

# SleepHrsNight vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = Depressed_bin, y = SleepHrsNight)) +
  geom_boxplot()

# Gender vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = Gender, fill = Depressed_bin)) +
  geom_bar(position = "fill")

# PhysActive vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = PhysActive, fill = Depressed_bin)) +
  geom_bar(position = "fill")

# MaritalStatus vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = MaritalStatus, fill = Depressed_bin)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle=45, hjust=1))

# PhysActive nach MaritalStatus
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), 
       aes(x = PhysActive, fill = Depressed_bin)) +
  geom_bar(position = "fill") +
  facet_wrap(~ MaritalStatus)

# BMI vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = Depressed_bin, y = BMI)) +
  geom_boxplot()

# Poverty vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = Depressed_bin, y = Poverty)) +
  geom_boxplot()

# HealthGen vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = HealthGen, fill = Depressed_bin)) +
  geom_bar(position = "fill")

# SleepTrouble vs Depressed
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), aes(x = SleepTrouble, fill = Depressed_bin)) +
  geom_bar(position = "fill")

# library(ggmosaic)
# ggplot(filter(nhanes_sub, !is.na(Depressed_bin))) +
#  geom_mosaic(aes(x = product(Gender, MaritalStatus), fill = Depressed_bin)) +
#  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# SleepTrouble nach Gender
ggplot(filter(nhanes_sub, !is.na(Depressed_bin)), 
       aes(x = SleepTrouble, fill = Depressed_bin)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Gender)

```

### Generalized Linear Model (family: Binomial)

#### First model (selection of variables)

Previous research has shown that depression is often associated with demographic, socioeconomic, social, health, and lifestyle factors. Based on these findings and after visual inspection of the graphs above, the following 9 variables are included in the first model:

-   **Demographic**: Age, Gender\
-   **Socioeconomic**: Poverty\
-   **Social**: MaritalStatus\
-   **Health-related**: BMI, HealthGen\
-   **Sleep / Activity**: SleepHrsNight, SleepTrouble, PhysActive

These choices are supported by findings in NHANES-based studies, where factors such as poverty, marital status, physical activity, sleep problems, BMI, and general health perception were among the strongest predictors of depression (see e.g. Zhang et al., *BMC Medical Informatics and Decision Making*, 2025).

```{r}
#| label: "GLM_Binomial_1"

Depressed_data <- nhanes_sub %>%
  filter(!is.na(Depressed_bin))

glm_depressed_1 <- glm(Depressed_bin ~ Age + Gender + Poverty + MaritalStatus + BMI +
                   HealthGen + SleepHrsNight + SleepTrouble + PhysActive,
                   data = Depressed_data,
                   family = binomial)
summary(glm_depressed_1)


```

#### Handling of missings

703 observation were removed as **missing**. In which variable are they missing?

```{r}
#| label: "GLM_Binomial_missings"

vars_model <- c("Depressed_bin", "Age", "Gender", "Poverty", "MaritalStatus", "BMI", "HealthGen", "SleepHrsNight", "SleepTrouble", "PhysActive")

Depressed_data %>%
  select(all_of(vars_model)) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(),
                      names_to = "Variable", values_to = "Missing_Count")



```

It is feasible to **replace missings** in `Poverty`, `BMI` and `SleepHrsNight` by the **median**. For `MaritalStatus` I don't see a convincing approach to replace missings.

```{r}
#| label: "Impute_Missings"

Depressed_data_imp <- Depressed_data %>%
  mutate(
    Poverty = ifelse(is.na(Poverty), median(Poverty, na.rm = TRUE), Poverty),
    BMI = ifelse(is.na(BMI), median(BMI, na.rm = TRUE), BMI),
    SleepHrsNight = ifelse(is.na(SleepHrsNight), median(SleepHrsNight, na.rm = TRUE), SleepHrsNight),)

Depressed_data_imp %>%
  select(all_of(vars_model)) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(),
                      names_to = "Variable", values_to = "Missing_Count")

```

I will continue the modelling with this dataset where only `MaritalStatus` is missing for 236 cases.

```{r}
#| label: "GLM_Binomial_1b"

glm_depressed_1b <- glm(Depressed_bin ~ Age + Gender + Poverty + MaritalStatus + BMI +
                   HealthGen + SleepHrsNight + SleepTrouble + PhysActive,
                   data = Depressed_data_imp,
                   family = binomial)
summary(glm_depressed_1b)

```

#### First interpretation

There is **strong evidence against the null hypothesis** that `Gender`, `Poverty`, `MaritalStatus`, `HealthGen`, `SleepHrsNight`, `SleepTrouble` and `PhysActivitie` have no effect. Whereas `Age` and `BMI` show **no significant effect** on Depressed_bin.

#### Checking for Collinearity

I am concerned about potential collinearity between the predictors, as this could hinder a proper interpretation of their individual effects. Therefore, I checked the model for signs of collinearity.

```{r}
#| label: "GLM_Binomial_1b_Collinear"

vif(glm_depressed_1b)   # auf dein Modell anwenden


```

All VIF values are \<2, meaning the model does not suffer from collinearity.

#### Second model (Age, BMI removed)

```{r}
#| label: "GLM_Binomial_2"

glm_depressed_2 <- glm(Depressed_bin ~ Gender + Poverty + MaritalStatus + HealthGen + SleepHrsNight + SleepTrouble + PhysActive,
                       data = Depressed_data_imp,
                       family = binomial)
summary(glm_depressed_2)

```

Do all variables **contribute significantly** to the quality of the model?

```{r}
#| label: "Glm_depressed_2_drop1"

drop1(glm_depressed_2, test = "Chisq")

```

As expected, all 7 variables are highly relevant for the quality of the model

#### Third model: Conceptual collinearity

I think that the variable **`HealthGen`** is **conceptually collinear** with `Depressed_bin`, as self-rated general health strongly overlaps with depressive symptoms. Since the focus of this analysis is on understanding underlying relationships rather than maximizing predictive accuracy, **`HealthGen`** will be excluded from the final model.

```{r}
#| label: "GLM_Binomial_3"

glm_depressed_3 <- glm(Depressed_bin ~ Gender + Poverty + MaritalStatus + SleepHrsNight + SleepTrouble + PhysActive,
                       data = Depressed_data_imp,
                       family = binomial)
summary(glm_depressed_3)

```

#### Comparison 2nd and 3rd model

The 2nd model has Residual deviance of 5948.8, the 3rd of 6095.2. Therefore **both differ clearly from the Null deviance** of 6688.1. The 2nd model including `HealthGen` is slightly better, also while comparing AIC (5978.8 vs. 6117.2). However, **I prefer the third model**, as `HealthGen` is not causally independent from `Depressed_bin` and may reflect overlapping constructs rather than a distinct explanatory factor.

To evaluate the discriminatory power of the logistic regression model (3rd), a **ROC (Receiver Operating Characteristic)** analysis is performed using the pROC package.

```{r}
#| label: "GLM_Binomial_3_ROC"
#| message: false

roc(glm_depressed_3$y, fitted(glm_depressed_3))

```

The resulting **AUC of 0.71** indicates that the model distinguishes **reasonably well** between depressed and non-depressed individuals. While not exceptionally high, this level of accuracy is typical for psycho-social data and supports the model’s usefulness for explanatory purposes.

### Generalized Additive Model (family: Binomial)

Taking non-linear effects into account could improve the model. Therefore I check first visually if there might be non-linear relations between `Depressed_bin` and the variables `Poverty` and `SleepHrsNight`.

#### Exploratory plots

```{r}
#| label: "Plot_nonlinear_effects"
#| message: false
#| warning: false

ggplot(Depressed_data_imp, aes(x = Poverty, y = as.numeric(Depressed_bin) - 1)) +
  geom_jitter(height = 0.05, alpha = 0.3, color = "gray60") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "blue", se = FALSE) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(
    y = "Probability of Depression",
    title = "Poverty vs. Probability of Depression",
    subtitle = "Blue = logistic fit (GLM), Red = flexible smoother (LOESS)"
  )

ggplot(Depressed_data_imp, aes(x = SleepHrsNight, y = as.numeric(Depressed_bin) - 1)) +
  geom_jitter(height = 0.05, alpha = 0.3, color = "gray60") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "blue", se = FALSE) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(
    y = "Probability of Depression",
    title = "Sleep hours per night vs. Probability of Depression",
    subtitle = "Blue = logistic fit (GLM), Red = flexible smoother (LOESS)"
  )


```

The exploratory plots suggest that the relationship between `Poverty` and depression is not strictly linear, while `SleepHrsNight` shows a clear non-linear, U-shaped association with depression probability.

#### Adding smooth splines to the model

To capture these **non-linear patterns** more accurately, a Generalized Additive Model (**GAM**) with **smooth spline terms** for `Poverty` and `SleepHrsNight` is being estimated using the `mgcv` package.

```{r}
#| label: "GAM_Binomial_1"
#| message: false

gam_depressed_1 <- gam(Depressed_bin ~ s(Poverty) + s(SleepHrsNight) +
                       Gender + MaritalStatus + SleepTrouble + PhysActive,
                     data = Depressed_data_imp, family = binomial)
summary(gam_depressed_1)

```

The GAM results confirm that both `Poverty` and `SleepHrsNight` exhibit **significant non-linear effects** on depression probability, as indicated by high estimated degrees of freedom (**edf \> 6**) and p-values \< 0.001.

#### Final interpretation of the GAM model

We have strong evidence against the null hypothesis that several factors have no effect on depression. Being **male** (Estimate = –0.31, p \< 0.001), being **married** (–0.67, p \< 0.001), and being **physically active** (–0.38, p \< 0.001) are all associated with lower odds of depression, while experiencing **sleep trouble** (+0.85, p \< 0.001) is associated with substantially higher odds. Individuals who are **widowed** also show a significantly lower risk (–0.54, p \< 0.001) compared to the reference category.

The smooth terms for **Poverty** (edf = 6.29, p \< 0.001) and **SleepHrsNight** (edf = 7.84, p \< 0.001) further indicate **significant non-linear effects** on depression probability. Overall, the model explains about **10% of the deviance**, suggesting that both socioeconomic and lifestyle factors meaningfully contribute to the likelihood of depression.

### Measure of Fit

#### ANOVA: comparing GLM and GAM models

As a **preliminary check**, an analysis of deviance (ANOVA) is being conducted to test whether the GAM provides a statistically significant improvement over the simpler GLM.

```{r}
#| label: "ANOVA_GLM_GAM"
#| message: false

anova(glm_depressed_3, gam_depressed_1, test = "Chisq")

```

The analysis of deviance confirmed that **the GAM significantly improved model fit** compared to the linear model (χ² = 69.10, df = 12.14, p \< 0.001), indicating that the inclusion of non-linear smooth terms for `Poverty` and `SleepHrsNight` provides a statistically better representation of the data.

#### Akaike Information Criterion (AIC)

The AIC is a commonly used measure of model fit that balances model accuracy and complexity. It is chosen because it allows a direct **comparison between models with different numbers of parameters**, penalizing excessive complexity while rewarding better data fit.

```{r}
#| label: "AIC_GLM_GAM"
#| message: false

AIC(glm_depressed_3, gam_depressed_1)

```

The **GAM achieved a lower AIC** (6072.4) compared to the linear GLM (6117.2), indicating a **better overall model fit** even after accounting for the increased model complexity.
